---
title: "Planned Curtailment Deanonymization"
format: html
execute:
  cache: true
---

## Strategy

CAISO's Day-Ahead Market (DAM) bid data anonymizes generators using a numeric `RESOURCEBID_SEQ` identifier, concealing which physical resource is behind each bid curve. However, the **Prior Trade Day Outage Report** publishes named resources (`RESOURCE ID`) along with their curtailment windows and curtailed MW.

The core idea is:

> When a generator files a **planned partial curtailment**, it knows its reduced operating limit before the DAM closes. Its day-ahead bid curve should therefore reflect a maximum MW of approximately `RESOURCE PMAX MW − CURTAILMENT MW` — the expected available capacity. If we find an anonymous `RESOURCEBID_SEQ` whose daily max bid MW closely matches this expected available capacity on the same date, that sequence ID is a candidate match for the named resource.

**Matching logic:**

1. Filter outages to `PLANNED` + partial curtailments (`CURTAILMENT MW < RESOURCE PMAX MW`)
2. Compute `expected_available_mw = RESOURCE PMAX MW − CURTAILMENT MW` per resource per curtailment date
3. Compute `daily_max_mw` per `RESOURCEBID_SEQ` from the DAM bid data
4. **Pre-filter** sequence IDs using `ID_attributes.parquet`: only consider IDs whose all-time `max_mw` is within 10% of `RESOURCE PMAX MW`. This eliminates implausible size mismatches (e.g. a 500 MW resource matching against a 9 MW curtailment target) before the expensive join
5. Join on date — every size-compatible sequence ID active on a curtailment day becomes a candidate
6. Filter to candidates whose `daily_max_mw` is within `TOLERANCE` of `expected_available_mw`
7. **Score** each `(RESOURCE ID, RESOURCEBID_SEQ)` pair using a combined metric that rewards both more matching days *and* lower MW error — described below
8. Compute a **confidence ratio** comparing the top candidate's score to the runner-up, to flag ambiguous matches
9. **Resolve collisions** where the same `RESOURCEBID_SEQ` is claimed by multiple named resources — keep only the highest-confidence claim

**Why planned/DAM?** Planned curtailments are filed in advance and published on the prior trade day report before the DAM closes, so the generator has every reason to reflect the derating in their day-ahead bids. This makes the signal clean and predictable, unlike forced outages which may not appear in DAM bids at all.

---

### Scoring

The raw count of matching days (`n_matches`) alone is insufficient as a ranking
signal because it ignores *how closely* a sequence ID's bids track the expected
available capacity on those days. Two candidates might both match 71 days, but
one always bids within 0.1% of the target while the other consistently sits at
4.9% — the first is a much stronger match.

The combined score is:

$$\text{score} = n\_matches \times (1 - \bar{e})$$

where $\bar{e}$ is the mean absolute percentage error across all matching days.
This penalises candidates with high average MW error, so a candidate with 71
matches at 0% error outscores one with 71 matches at 4% error.

### Confidence

The **confidence ratio** is defined as:

$$\text{confidence} = \frac{\text{score}_{\text{top}}}{\text{score}_{\text{runner-up}}}$$

A ratio much greater than 1 means the top candidate is clearly better than the
next-best option — a strong deanonymization signal. A ratio near 1 indicates
the match is ambiguous and should not be trusted without additional corroborating
evidence.

When a named resource has **only one candidate** after filtering (no runner-up
exists), the confidence ratio is set to `Inf` — it is an unambiguous match by
default, since no other sequence ID is size-compatible and temporally consistent
with this resource's curtailment history. This is assigned explicitly using
`if_else(is.na(runner_up), Inf, score / runner_up)` rather than dividing by a
small epsilon, which would produce arbitrarily large but finite values and
contaminate the distribution.

### Collision resolution

A given `RESOURCEBID_SEQ` can only correspond to one physical generator. If
multiple named resources claim the same sequence ID, we keep only the claim with
the highest confidence ratio and discard the rest.

---

## Setup

```{r}
#| label: setup
#| message: false
#| cache: false
library(duckdb)
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
```

```{r}
#| label: load-data
#| message: false

con <- dbConnect(duckdb())
tbl_pointer <- tbl(con, "read_parquet('/Users/sumanth/Desktop/BidCurves/CAISO_dam_big.parquet')")

# DAM parquet has already been pre-filtered to EN generator energy bids
data <- tbl_pointer |>
  select(RESOURCEBID_SEQ, STARTTIME, SCH_BID_XAXISDATA, SELFSCHEDMW) |>
  collect()

outages  <- read_parquet(here("prior_trade_day_outages.parquet"))
id_attrs <- read_parquet(here("ID_attributes.parquet")) |>
  select(RESOURCEBID_SEQ, max_mw)
```

---

## Step 1: Planned Partial Curtailments

Filter to planned outages where the generator is derated but not fully offline
(`CURTAILMENT MW < RESOURCE PMAX MW`), and deduplicate rows that appear across
multiple report publication dates for the same event.

```{r}
#| label: planned-partial

planned_partial <- outages |>
  filter(
    `OUTAGE TYPE` == "PLANNED",
    `CURTAILMENT MW` < `RESOURCE PMAX MW`
  ) |>
  mutate(
    curtailment_date      = as.Date(`CURTAILMENT START DATE TIME`),
    expected_available_mw = `RESOURCE PMAX MW` - `CURTAILMENT MW`
  ) |>
  distinct(
    `RESOURCE ID`, `RESOURCE NAME`, curtailment_date,
    `RESOURCE PMAX MW`, `CURTAILMENT MW`, expected_available_mw
  )

cat("Unique planned partial curtailment events:", nrow(planned_partial), "\n")
cat("Unique named resources:                   ", n_distinct(planned_partial$`RESOURCE ID`), "\n")
```

---

## Step 2: Daily Max MW from DAM

Compute the maximum MW offered by each anonymous sequence ID on each calendar
day. `pmax()` handles the element-wise maximum across the economic bid column
and the self-schedule column, ignoring NAs.

```{r}
#| label: daily-max
#| warning: false

daily_max <- data |>
  mutate(
    date       = as.Date(STARTTIME),
    row_max_mw = pmax(SCH_BID_XAXISDATA, SELFSCHEDMW, na.rm = TRUE)
  ) |>
  summarise(
    daily_max_mw = max(row_max_mw, na.rm = TRUE),
    .by = c(RESOURCEBID_SEQ, date)
  ) |>
  filter(is.finite(daily_max_mw))

cat("Sequence-ID / date combinations:", nrow(daily_max), "\n")
```

---

## Step 3: Pre-filter by PMAX Compatibility

Before joining on date, restrict sequence IDs to those whose all-time observed
`max_mw` (from `ID_attributes.parquet`) is within 10% of `RESOURCE PMAX MW`
for at least one curtailment event. This eliminates implausible size mismatches
before the expensive cross-join.

```{r}
#| label: pmax-prefilter

PMAX_TOLERANCE <- 0.10

pmax_lookup <- planned_partial |>
  distinct(`RESOURCE ID`, `RESOURCE PMAX MW`) |>
  cross_join(id_attrs) |>
  filter(
    abs(max_mw - `RESOURCE PMAX MW`) / `RESOURCE PMAX MW` <= PMAX_TOLERANCE
  ) |>
  select(`RESOURCE ID`, RESOURCEBID_SEQ)

cat("Compatible (resource, seq ID) pairs after PMAX filter:", nrow(pmax_lookup), "\n")

daily_max_filtered <- daily_max |>
  semi_join(pmax_lookup |> distinct(RESOURCEBID_SEQ), by = "RESOURCEBID_SEQ")

cat("Daily max rows after PMAX filter:", nrow(daily_max_filtered), "\n")
```

---

## Step 4: Match on Date + Capacity

Join curtailment events to size-compatible sequence IDs active on the same
date, then keep only those whose `daily_max_mw` falls within `TOLERANCE` of
`expected_available_mw`.

```{r}
#| label: match

TOLERANCE <- 0.05  # ±5% of expected available MW

matches <- planned_partial |>
  inner_join(pmax_lookup, by = "RESOURCE ID", relationship = "many-to-many") |>
  inner_join(daily_max_filtered,
             by = join_by(curtailment_date == date, RESOURCEBID_SEQ),
             relationship = "many-to-many") |>
  filter(
    expected_available_mw > 0,
    abs(daily_max_mw - expected_available_mw) / expected_available_mw <= TOLERANCE
  ) |>
  select(
    `RESOURCE ID`, `RESOURCE NAME`, curtailment_date,
    `RESOURCE PMAX MW`, `CURTAILMENT MW`, expected_available_mw,
    RESOURCEBID_SEQ, daily_max_mw
  )

cat("Total candidate match rows:", nrow(matches), "\n")
```

---

## Step 5: Score Candidate Matches

For each `(RESOURCE ID, RESOURCEBID_SEQ)` pair, compute:

- `n_matches` — number of curtailment days where the daily max MW was within tolerance
- `mean_pct_err` — mean absolute % error across those matching days
- `score = n_matches × (1 − mean_pct_err)` — combined metric rewarding both frequency and MW closeness

```{r}
#| label: scores

match_scores <- matches |>
  mutate(
    abs_pct_err = abs(daily_max_mw - expected_available_mw) / expected_available_mw
  ) |>
  summarise(
    n_matches    = n(),
    mean_pct_err = mean(abs_pct_err),
    score        = n_matches * (1 - mean_pct_err),
    .by = c(`RESOURCE ID`, `RESOURCE NAME`, RESOURCEBID_SEQ)
  ) |>
  arrange(desc(score))

match_scores
```

---

## Step 6: Top Candidates with Confidence

For each named resource, select the single best-scoring sequence ID. The
runner-up score is extracted with `nth(..., 2)`. When only one candidate exists
`nth()` returns `NA` — in that case confidence is set explicitly to `Inf` via
`if_else()`, correctly reflecting an unambiguous match. This avoids dividing by
a small epsilon which would produce arbitrarily large finite values that
misrepresent the distribution.

```{r}
#| label: top-candidates

top_candidates <- match_scores |>
  filter(n_matches >= 2) |>
  mutate(
    rank       = rank(-score, ties.method = "first"),
    runner_up  = nth(sort(score, decreasing = TRUE), 2),
    confidence = if_else(is.na(runner_up), Inf, score / runner_up),
    .by = `RESOURCE ID`
  ) |>
  filter(rank == 1) |>
  select(-rank, -runner_up) |>
  arrange(desc(confidence))

top_candidates
```

---

## Step 7: Resolve Seq ID Collisions

A `RESOURCEBID_SEQ` can only correspond to one physical generator. Where
multiple named resources claim the same sequence ID, retain only the
highest-confidence claim.

```{r}
#| label: resolve-collisions

collisions <- top_candidates |>
  filter(n() > 1, .by = RESOURCEBID_SEQ) |>
  arrange(RESOURCEBID_SEQ, desc(confidence))

cat("Seq IDs claimed by multiple resources:", n_distinct(collisions$RESOURCEBID_SEQ), "\n")
cat("Resources dropped in collision resolution:",
    nrow(collisions) - n_distinct(collisions$RESOURCEBID_SEQ), "\n")

top_candidates_clean <- top_candidates |>
  slice_max(confidence, n = 1, with_ties = FALSE, by = RESOURCEBID_SEQ)

cat("Final unique matches:", nrow(top_candidates_clean), "\n")
```

---

## Step 8: Inspect Match Detail

For each final candidate, examine the per-day MW error.

```{r}
#| label: match-detail

match_detail <- matches |>
  semi_join(top_candidates_clean, by = c("RESOURCE ID", "RESOURCEBID_SEQ")) |>
  mutate(
    mw_error  = daily_max_mw - expected_available_mw,
    pct_error = mw_error / expected_available_mw
  )

match_detail |>
  select(`RESOURCE ID`, RESOURCEBID_SEQ, curtailment_date,
         expected_available_mw, daily_max_mw, pct_error) |>
  arrange(`RESOURCE ID`, curtailment_date)
```


## Confidence Distribution

Visualise the spread of confidence ratios across all final candidates. `Inf`
values (single-candidate resources) are excluded from the histogram since they
have no runner-up by definition, and are reported separately. The x-axis is
log-scaled to handle the wide range of finite confidence values. The dashed line
marks a ratio of 2 — a practical threshold for considering a match trustworthy.

```{r}
#| label: plot-confidence
#| fig-width: 8
#| fig-height: 4

plot_data <- top_candidates_clean |>
  filter(is.finite(confidence))

n_inf <- sum(!is.finite(top_candidates_clean$confidence))

ggplot(plot_data, aes(x = confidence)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "firebrick") +
  scale_x_log10(labels = scales::comma) +
  annotate("text", x = 2.2, y = Inf, label = "confidence > 2",
           hjust = 0, vjust = 1.5, color = "firebrick", size = 3.5) +
  annotate("text", x = Inf, y = Inf,
           label = paste0(n_inf, " resources with Inf confidence\n(no runner-up candidate)"),
           hjust = 1.05, vjust = 1.5, size = 3.5, color = "grey40") +
  labs(
    title = "Distribution of Confidence Ratios (log scale)",
    subtitle = "Top candidate score / runner-up score, per named resource (finite values only)",
    x = "Confidence Ratio (log scale)",
    y = "Count"
  ) +
  theme_minimal()
```

---

## Summary

```{r}
#| label: summary

high_conf <- sum(top_candidates_clean$confidence > 2, na.rm = TRUE)
inf_conf  <- sum(!is.finite(top_candidates_clean$confidence))

cat("Planned partial curtailment events:          ", nrow(planned_partial), "\n")
cat("Candidate (resource, seq) pairs — any day:   ", nrow(match_scores), "\n")
cat("Strong candidates (\u22652 matching days):         ", sum(match_scores$n_matches >= 2), "\n")
cat("Top candidates before collision resolution:  ", nrow(top_candidates), "\n")
cat("Final unique matches after resolution:       ", nrow(top_candidates_clean), "\n")
cat("  of which Inf confidence (no runner-up):    ", inf_conf, "\n")
cat("  of which finite confidence > 2:            ", high_conf, "\n")
cat("Unique named resources matched:              ", n_distinct(top_candidates_clean$`RESOURCE ID`), "\n")
cat("Unique RESOURCEBID_SEQ matched:              ", n_distinct(top_candidates_clean$RESOURCEBID_SEQ), "\n")
```

---

## Export Results

```{r}
#| label: export-csv
#| message: false

# 1. Final deanonymization map: one row per (named resource → seq ID) pair
#    Joined with resource nameplate PMAX and the seq ID's all-time observed max MW
# Some resources have multiple PMAX values in planned_partial (nameplate changed
# over time). Take the maximum to avoid duplicate rows in the join.
resource_pmax <- planned_partial |>
  summarise(`RESOURCE PMAX MW` = max(`RESOURCE PMAX MW`), .by = `RESOURCE ID`)

final_with_pmax <- top_candidates_clean |>
  left_join(resource_pmax, by = "RESOURCE ID") |>
  left_join(id_attrs |> rename(seq_max_mw = max_mw), by = "RESOURCEBID_SEQ") |>
  select(
    `RESOURCE ID`, `RESOURCE NAME`, `RESOURCE PMAX MW`,
    RESOURCEBID_SEQ, seq_max_mw,
    n_matches, mean_pct_err, score, confidence
  )

write.csv(
  final_with_pmax,
  here("curtailment_matches_final.csv"),
  row.names = FALSE
)

# 2. Per-day match detail for every final candidate
write.csv(
  match_detail |>
    select(`RESOURCE ID`, RESOURCEBID_SEQ, curtailment_date,
           expected_available_mw, daily_max_mw, mw_error, pct_error) |>
    arrange(`RESOURCE ID`, curtailment_date),
  here("curtailment_match_detail.csv"),
  row.names = FALSE
)

# 3. All candidate scores (before collision resolution) — useful for auditing
write.csv(
  match_scores,
  here("curtailment_match_scores_all.csv"),
  row.names = FALSE
)

cat("Wrote curtailment_matches_final.csv       —", nrow(final_with_pmax), "rows\n")
cat("Wrote curtailment_match_detail.csv        —", nrow(match_detail), "rows\n")
cat("Wrote curtailment_match_scores_all.csv    —", nrow(match_scores), "rows\n")
```