---
title: "Bid Curve Shenanigans"
subtitle: ""
format: 
  dashboard:
    orientation: rows
    theme: flatly
server: shiny
embed-resources: true
---

```{r}
#| label: load-packages
#| include: false
#| context: setup

library(tidyverse)
library(plotly)
library(duckdb)
library(gt)
library(Matrix)
```

```{r}
#| label: load-data
#| include: false
#| context: setup

con <- dbConnect(duckdb())
tbl_pointer <- tbl(con, "read_parquet('CAISO_dam_big.parquet')")
# tbl_pointer <- tbl(con, "read_parquet('interp_dataset/id_952796.parquet')")
# tbl_pointer|>
#   head()|>
#   collect()
# average_bid_curve <- read_csv("average_bid_curve_test.csv")

interp_tbl <- arrow::open_dataset("interp_dataset")

id_matches <- readRDS(here::here("cos_sim.rda"))

features <- arrow::read_parquet("feature_vectors.parquet")

gen_matches_all <- read_csv("curtailment_matches_final.csv")
```

```{r}
#| context: server
max_mw <- eventReactive(input$submit_id, {
    req(input$resource_id)

    tbl_pointer |>
    filter(RESOURCEBID_SEQ == input$resource_id) |>
    summarise(max_mw = max(SCH_BID_XAXISDATA, na.rm = TRUE)) |>
    collect() |>
    pull(max_mw)
})


average_bid_curve <- eventReactive(input$submit_id, {
  req(input$resource_id)

  max_mw2 <- max_mw()

  interp_tbl |>
    filter(RESOURCEBID_SEQ == input$resource_id,
           date(date) >= input$dates[1],
           date(date) <= input$dates[2]) |>
    # group by each hour each day
    group_by(hour, pct_grid, daily_x) |>
    summarise(
      avg_price = mean(price_interp, na.rm = TRUE),
      sd = sd(price_interp, na.rm = TRUE),
      # number of days included in the average
      n_days_used = sum(!is.na(price_interp)),
      .groups = "drop"
    ) |>
    # if no days included in avg. filter out
    filter(n_days_used != 0) |>
    collect() |>
    mutate(orig_MW = (pct_grid/100)*max_mw2)
    # mutate(orig_MW = pct_grid) |>
})

best_matches <- eventReactive(input$submit_id, {
    req(input$resource_id)

    tibble(
        matched_id = colnames(id_matches),
        similarity = as.numeric(id_matches[as.character(input$resource_id), ])
        ) |>
        filter(matched_id != as.character(input$resource_id)) |>
        slice_max(order_by = similarity, n = 5)
    })

gen_matches <- eventReactive(input$submit_id, {
    req(input$resource_id)

    gen_matches_all |>
    filter(RESOURCEBID_SEQ == input$resource_id) |>
    select(-RESOURCEBID_SEQ, -`RESOURCE ID`) |>
    rename(
        # "Resource ID" = RESOURCEBID_SEQ,
           "Resource Name" = `RESOURCE NAME`,
           "Resource Pmax" = `RESOURCE PMAX MW`,
           "ID Max MW" = seq_max_mw,
           "# of Matches" = n_matches,
           "Mean % Error" = mean_pct_err,
           "Score" = score,
           "Confidence" = confidence) |>
    pivot_longer(
        cols = `Resource Pmax`:Confidence,
        names_to = "Matching Attributes",
        values_to = "Value"
    ) |>
    mutate(Value = round(Value, 3),
           `Resource Name` = if_else(`Resource Name` == lag(`Resource Name`, default = ""), "", `Resource Name`)) 
})

matches_by_period <- reactive({

  target_id <- as.character(input$resource_id)

  target_dates <- features |>
    filter(RESOURCEBID_SEQ == target_id) |>
    summarise(
      target_start = first(start),
      target_end = first(end)
    )

  target_start <- target_dates$target_start
  target_end <- target_dates$target_end

  match_tbl <- best_matches()

  match_tbl |>
    left_join(
      features |>
        select(RESOURCEBID_SEQ, start, end),
      by = c("matched_id" = "RESOURCEBID_SEQ")
    ) |>
    mutate(
      period = case_when(
        end < target_start ~ "Before",
        start > target_end ~ "After",
        start <= target_end & end >= target_start ~ "During"
      )
    )
})
```

```{r}
#| label: prep-data
#| include: false
#| context: server

# # interpolate every 1% of max MW
# pct_grid <- tibble(pct_grid = seq(0, 100, by = 1))

# get maximum MW for id
# max_mw <- test |>
#   summarise(max_mw = max(SCH_BID_XAXISDATA, na.rm = TRUE)) |>
#   pull(max_mw)

# # get outliers for plot
# outliers <- tbl_pointer |>
#        mutate(SCH_BID_XAXISDATA = coalesce(as.double(SELFSCHEDMW), SCH_BID_XAXISDATA)) |>
#        filter(RESOURCEBID_SEQ == 952796) |>
#        select(RESOURCEBID_SEQ, SCH_BID_XAXISDATA, SCH_BID_Y1AXISDATA, STARTTIME) |>
#        mutate(date = as_datetime(STARTTIME),
#               hour = hour(date),
#               # gets the % of the max MW for each MW value in bids
#               mw_pct = (SCH_BID_XAXISDATA/max(SCH_BID_XAXISDATA, na.rm = TRUE))*100,
#               mw_pct_round = round(mw_pct)) |>
#       filter(!is.na(mw_pct_round), !is.na(SCH_BID_Y1AXISDATA)) |>
#       group_by(date, hour, mw_pct_round) |>
#       summarise(
#           SCH_BID_Y1AXISDATA = mean(SCH_BID_Y1AXISDATA),
#           .groups = "drop") |>
#       arrange(date, hour, mw_pct_round) |>
#       collect() |>
#       left_join(average_bid_curve, 
#                 by = join_by(hour,
#                 mw_pct_round == pct_grid)) |>
#       filter(SCH_BID_Y1AXISDATA > avg_price + 2*sd,
#             SCH_BID_Y1AXISDATA < avg_price - 2*sd)

# # interpolate values for pct_grid
# average_bid_curve <- test_pct |>
#   group_by(date, hour) |>
#   # do for each date, hour combination
#   group_modify(\(.x, .y) {
    
#     # actual bids in that date, hour
#     submitted_bids <- .x |>
#       select(mw_pct_round, SCH_BID_Y1AXISDATA) |>
#       arrange(mw_pct_round)
    
#     # min and max MW in bids as a pct - so we don't interpolate outside actual MWs in bids
#     min_pct <- min(submitted_bids$mw_pct_round, na.rm = TRUE)
#     max_pct <- max(submitted_bids$mw_pct_round, na.rm = TRUE)
    
#     pct_grid |>
#       mutate(
#         price_interp = map_dbl(pct_grid, \(g) {
          
#           # so we don't interpolate outside actual bids
#           if (g < min_pct || g > max_pct) {
#             return(NA)
#           }
          
#           # exact match: use actual submitted price
#           exact <- submitted_bids |>
#             filter(mw_pct_round == g)
          
#           if (nrow(exact) > 0) {
#             return(exact$SCH_BID_Y1AXISDATA[1])
#           }
          
#           # otherwise use closest actual bid with smaller percentage
#           closest_lower_mw <- submitted_bids |>
#             filter(mw_pct_round < g) |>
#             slice_max(mw_pct_round, n = 1, with_ties = FALSE)
          
#           if (nrow(closest_lower_mw) == 0) {
#             return(NA_real_)
#           }
          
#           closest_lower_mw$SCH_BID_Y1AXISDATA
#         })
#       )
#   }) |>
#   ungroup() |>
# # 24 hour bid curves
#   mutate(
#     # ex. 5:00 is 500
#     # if we wanted to see a bid in this hour at 20% MW, it'd be 520
#     daily_x = hour * 100 + pct_grid
#   ) |>
#   # group by each hour each day
#   group_by(hour, pct_grid, daily_x) |>
#   summarise(
#     avg_price = mean(price_interp, na.rm = TRUE),
#     sd = sd(price_interp, na.rm = TRUE),
#     # number of days included in the average
#     n_days_used = sum(!is.na(price_interp)),
#     .groups = "drop"
#   ) |>
#   # if no days included in avg. filter out
#   filter(n_days_used != 0)
```

# About

::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## CAISO Bid Curves

Here you can view the average bid curve for market participants in CAISO's Day-Ahead Market (DAM). Enter an ID to view its average bid curve and potential outliers in bidding patterns for January 2023 to June 2025 in the **Average Bid Curve** tab. Hover over plots to see more information.
:::


## Row
::: {.card width="20%"}
## Explore a Resource ID

To populate the Average Bid Curve, ID Matching, and Deanonymization with results, first submit the anonymized DAM ID you want to view results for.

```{r}
numericInput(
    inputId = "resource_id",
    label = "Please enter a Resource ID:",
    value = NULL
)

actionButton(
  inputId = "submit_id",
  label = "Submit ID",
  class = "btn-primary"
)

uiOutput("id_warning")
```

:::

::: {.card}

## **Navigate the dashboard**

<!-- <strong style="font-weight: bold; color: primary;">Average Bid Curve</strong>

<details><summary><strong>Calculating an Average Bid Curve</strong></summary>
To compare bid curves across days, each ID's MW values are rescaled to a common percentage scale and averaged across daily bidding patterns.

1. **Rescale MW values**  
   Convert each submitted MW quantity into a percentage of the resource’s maximum submitted MW.

2. **Round to a 1% grid**  
   Round each MW percentage to the nearest whole percent.

3. **Interpolate within each hourly curve**  
   For each day-hour curve, fill in prices on a 0–100% MW grid.

   - Use the actual submitted price when one exists.
   - Otherwise, use the closest lower submitted MW percentage.
   - Do not extrapolate beyond the observed MW range for that hour.

4. **Stack hourly curves into a daily curve**  
   Combine the 24 hourly curves into one continuous daily curve.

5. **Average across days**  
   Average prices across all days at each hour-percent combination to create the resource’s typical 24-hour bid curve.
</details>

<strong style="font-weight: bold; color: primary;">ID Matching</strong>

<strong style="font-weight: bold; color: primary;">Deanonymization</strong> -->

::: {.panel-tabset}

# Average Bid Curve

To compare bid curves across days, each ID's MW values are rescaled to a common percentage scale and averaged across daily bidding patterns.

1. **Rescale MW values**  
   Convert each submitted MW quantity into a percentage of the resource’s maximum submitted MW.

2. **Round to a 1% grid**  
   Round each MW percentage to the nearest whole percent.

3. **Interpolate within each hourly curve**  
   For each day-hour curve, fill in prices on a 0–100% MW grid.

   - Use the actual submitted price when one exists.
   - Otherwise, use the closest lower submitted MW percentage.
   - Do not extrapolate beyond the observed MW range for that hour.

4. **Stack hourly curves into a daily curve**  
   Combine the 24 hourly curves into one continuous daily curve.

5. **Average across days**  
   Average prices across all days at each hour-percent combination to create the resource’s typical 24-hour bid curve.

# ID Matching

Anonymous IDs usually do not always last throughout the lifetime of the dataset, suggesting that IDs get shuffled or changed for at least some plants. Following the assumption that Resources will have similar behavior throughout their lifetime, this pipeline aims to match deprecated IDs to new IDs that pop up after them, based on some features we determined were important. These include Max MW bid, Mean and median # bids submitted throughout the lifetime, Single-point bid frequency, Null bid percentage, self-scheduled percentage, the hours bid and negative bids. The output is an ID to ID match and similarity. 

# Deanonymization

When a generator files a planned partial curtailment, it knows its reduced operating limit before the DAM closes. Its day-ahead bid curve should therefore reflect a maximum MW of approximately its PMax - Curtailment MW, to give an expected available capacity. If we find an anonymous resource whose daily max bid MW closely matches this expected available capacity, then it is a candidate match for the named resource. Further filtering is based on the number of matched curtailment events and the closeness of the max bid to the resource’s pmax. A score and a confidence are given for the matches. The score is based on the mean absolute percentage error of bids across all matching days along with the number of matches. The confidence (0-1 scale) is based on how much the top score is better than the runner up. A score of 1 represents an unambiguous match.
:::


:::

# Deanonymization

::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## ID to Resource Name

This page displays the most similar resources for the selected ID using curtailment matching. Matches are ordered in terms of confidence. Hover over words highlighted in <span style="color:#325D88;">blue</span> for more information.

To view matches for a different ID, enter a new Resource ID on the **About** page and click **Submit ID**.
:::
<!-- 
he table on the left lists definitions of the attributes using in matching. The table on the right displays the matched resource along with matching scores and information about the resource used in the matching process. -->

## Row

<!-- ### Column  -->

<!-- ::: {.card}

## Matching Attribute Definitions

- **Resource Name:** Official name of the matched resource.
- **Resource Pmax:** CAISO reported Physical Maximum output for the matched resource.
- **ID Max MW:** Anonymized ID’s all time maximum MW bid
- **# of Matches:** Number of matched curtailment events.
- **Score:** Number of Matches x (1 - mean absolute percentage error across all matching days)
- **Confidence:** 0-1 scale of how much better the top score compared to the runner up (1 being an unambiguous match)

::: -->

### Column

```{r}
# gt_output("gen_match_tbl")
uiOutput("gen_match_summary")
```

```{r}
#| context: server

# output$gen_match_summary <- renderUI({

#   req(input$resource_id)

#   target_id <- as.character(input$resource_id)

#   gen_row <- gen_matches_all |>
#     filter(RESOURCEBID_SEQ == input$resource_id) |>
#     slice(1)

#   validate(
#     need(nrow(gen_row) > 0, "No deanonymization match found for this ID.")
#   )

#   resource_name <- gen_row$`RESOURCE NAME`
#   resource_pmax <- round(gen_row$`RESOURCE PMAX MW`, 3)
#   id_max_mw <- round(gen_row$seq_max_mw, 3)
#   n_matches <- gen_row$n_matches
#   score <- round(gen_row$score, 3)
#   confidence <- round(gen_row$confidence, 3)

#   HTML(glue::glue(
#     '
#     <div class="card" style="padding: 22px; font-size: 18px; line-height: 1.7;">
#       <p>
#         The most likely resource for anonymous ID <strong>{target_id}</strong> is 
#         <strong style="color#18BC9C;">{resource_name}</strong>.
#       </p>

#       <p>
#         Our final score for this match is <strong style="color:#3498DB;">{score}</strong>, with a confidence of 
#         <strong style="color:#F39C12;">{confidence}</strong>.
#       </p>

#       <p>
#         Why? The resource Pmax is <strong style="color:#a2c7ea;">{resource_pmax} MW</strong>, and ID {target_id}’s
#         maximum MW ever submitted in a bid is <strong style="color:#a2c7ea;">{id_max_mw} MW</strong>.
#         There were <strong>{n_matches}</strong> curtailment events that matched this ID.
#       </p>

#     </div>
#     '
#   ))
# })

output$gen_match_summary <- renderUI({

  req(input$resource_id)

  target_id <- as.character(input$resource_id)

  gen_row <- gen_matches_all |>
    filter(RESOURCEBID_SEQ == input$resource_id) |>
    slice(1)

  resource_name <- gen_row$`RESOURCE NAME`
  resource_pmax <- round(gen_row$`RESOURCE PMAX MW`, 3)
  id_max_mw <- round(gen_row$seq_max_mw, 3)
  n_matches <- gen_row$n_matches
  score <- round(gen_row$score, 3)
  confidence <- round(gen_row$confidence, 3)

  div(
    class = "card",
    style = "padding: 22px; font-size: 18px; line-height: 1.7;",

    p(
      "The most likely ",
    #   tags$span(
    #     "resource ⓘ",
    #     title = "Official name of the matched resource.",
    #     style = "cursor: help;"
    #   ),
      tags$span(
        "resource",
        title = "Official name of the matched resource.",
        style = "
            color: #325D88;
            cursor: help;
        "
        ),
      " for anonymous ID ",
      tags$strong(target_id),
      " is ",
      tags$strong(
        resource_name,
        style = "color:#18BC9C;"
      ),
      "."
    ),

    p(
      "Our final ",
    #   tags$span(
    #     "score ⓘ",
    #     title = "Number of matches multiplied by (1 - the mean absolute percentage error).",
    #     style = "cursor: help;"
    #   ),
      tags$span(
        "score",
        title = "Number of matches multiplied by (1 - the mean absolute percentage error).",
        style = "
            color: #325D88;
            cursor: help;"
        ),
      " for this match is ",
      tags$strong(score, style = "color:#3498DB;"),
      ", with a ",
    #   tags$span(
    #     "confidence ⓘ",
    #     title = "How much better the top score is compared to the runner-up. Values closer to 1 indicate a clearer match.",
    #     style = "cursor: help;"
    #   ),
      tags$span(
        "confidence",
        title = "How much better the top score is compared to the runner-up. Values closer to 1 indicate a clearer match.",
        style = "
            color: #325D88;
            cursor: help;"
        ),
      " of ",
      tags$strong(confidence, style = "color:#F39C12;"),
      "."
    ),

    p(
      "Why? The resource",
    #   tags$span(
    #     "Pmax ⓘ",
    #     title = "CAISO reported physical maximum output for the matched resource.",
    #     style = "cursor: help;"
    #   ),
      tags$span(
        "Pmax",
        title = "CAISO reported physical maximum output for the matched resource.",
        style = "
            color: #325D88;
            cursor: help;"
        ),
      " is ",
      tags$strong(paste0(resource_pmax, " MW"), style = "color:#E74C3C;"),
      ", and ID ",
      target_id,
      "'s maximum MW ever submitted in a bid is ",
      tags$strong(paste0(id_max_mw, " MW"), style = "color:#E74C3C;"),
      ". There were ",
      tags$strong(n_matches),
      " curtailment events that matched this ID."
    )
  )
})
```


# ID Matching

::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}

## ID to ID Matching

This page displays the most similar Resource IDs for the selected ID. The table on the left lists the top matching IDs ranked by similarity score. Similarity scores range from 0 to 1, with values closer to 1 indicating more similar bidding behavior. The table on the right displays the attributes used in the matching process, where you can compare the characteristics of the matched resources.

To view matches for a different resource, enter a new Resource ID on the **About** page and click **Submit ID**.
:::


## Row

### Column {width="26%"}

::: {.card}

## Attribute Definitions

- **Mean knots:** The mean number of bids submitted in a curve.
- **Median knots:** The median number of bids submitted in a curve.
- **Single-point curves:** The percentage of bids that are single-point curves.
- **Null bids:** Whether NULL bids are ever submitted.
- **Negative bids:** Whether the ID submits negative bids.
- **Self-scheduled bids:** The percentage of bids that are self-scheduled.
- **Max MW:** The maximum MW quantity ever submitted in a bid.

:::

### Column {width="24%"}

```{r}
gt_output("id_match_tbl")
```

```{r}
#| context: server

output$id_match_tbl <- render_gt({

    target_id <- as.character(input$resource_id)

    best_matches_tbl <- best_matches()

    best_matches_tbl |>
        gt() |>
        cols_label(
            matched_id = md("**Resource ID**"),
            similarity = md("**Similarity**")
            ) |>
            tab_header(
                title = glue::glue("ID Matches for {target_id}")
                ) |>
            tab_style(
                style = list(cell_fill(color = "#325D88"),
                            cell_text(color = "white")),
                        locations = cells_title(groups = "title")
            ) |>
            tab_style(
                style = cell_borders(sides = "right",
                                    color = "lightgray",
                                    weight = px(2)),
                locations = list(cells_body(columns = matched_id),
                                cells_column_labels(columns = matched_id))
            ) |>
            cols_align(align = "center")
})
```

### Column {width="50%"}

```{r}
gt_output("id_features_tbl")
```
```{r}
#| context: server

output$id_features_tbl <- render_gt({

    target_id <- as.character(input$resource_id)

    best_matches_tbl <- best_matches()

    features |>
    filter(RESOURCEBID_SEQ %in% best_matches_tbl$matched_id) |>
        pivot_wider(
            names_from = RESOURCEBID_SEQ,
            values_from = Value
        ) |>
        select(
            Feature,
            all_of(best_matches_tbl$matched_id)
            ) |>
        gt() |>
        cols_label(
            Feature = md("**Attribute**")
            ) |>
        tab_spanner(
            label = md("**Resource ID**"),
            columns = -Feature
        )  |>
        tab_header(
                title = glue::glue("Attributes of Top Matches for {target_id}")
                )
})
```

# TEST PAGE
```{r}
uiOutput("id_match_words")
```

```{r}
#| context: server

get_feature <- function(data, id, feature_name) {
  data |>
    filter(
      RESOURCEBID_SEQ == id,
      Feature == feature_name
    ) |>
    pull(Value) |>
    first()
}

has_text <- function(value, id, label) {
  if (value %in% c("Yes", "TRUE", TRUE, 1, "1")) {
    glue::glue("{id} has submitted {label}.")
  } else {
    glue::glue("{id} has never submitted {label}.")
  }
}

output$id_match_words <- renderUI({

  req(input$resource_id)

  target_id <- as.character(input$resource_id)
  best_matches_tbl <- best_matches()

  validate(
    need(nrow(best_matches_tbl) > 0, "No ID matches found for this Resource ID.")
  )

  # target ID features
  target_max_mw <- get_feature(features, target_id, "Max MW")
  target_mean <- get_feature(features, target_id, "Mean knots")
  target_median <- get_feature(features, target_id, "Median knots")
  target_single <- get_feature(features, target_id, "Single-point curves")
  target_self <- get_feature(features, target_id, "Self-scheduled bids")
  target_null <- get_feature(features, target_id, "Null bids")
  target_negative <- get_feature(features, target_id, "Negative bids")

  target_null_sentence <- has_text(target_null, target_id, "NULL bids")
  target_negative_sentence <- has_text(target_negative, target_id, "negative bids")

  # make one paragraph per match
  match_paragraphs <- purrr::map(seq_len(nrow(best_matches_tbl)), function(i) {

    match_id <- best_matches_tbl$matched_id[i]
    sim <- round(best_matches_tbl$similarity[i], 3)

    match_max_mw <- get_feature(features, match_id, "Max MW")
    match_mean <- get_feature(features, match_id, "Mean knots")
    match_median <- get_feature(features, match_id, "Median knots")
    match_single <- get_feature(features, match_id, "Single-point curves")
    match_self <- get_feature(features, match_id, "Self-scheduled bids")
    match_null <- get_feature(features, match_id, "Null bids")
    match_negative <- get_feature(features, match_id, "Negative bids")

    match_null_sentence <- has_text(match_null, match_id, "NULL bids")
    match_negative_sentence <- has_text(match_negative, match_id, "negative bids")

    tags$p(
      tags$strong(glue::glue("Match {i}: ")),
      glue::glue("ID {match_id} has a similarity of "),
      tags$strong(style = "color:#325D88;", sim),
      glue::glue(". This ID’s maximum MW submitted in a bid is "),
      tags$strong(match_max_mw),
      glue::glue(" MW. ID {match_id} submits on average "),
      tags$strong(match_mean),
      glue::glue(" bids in a curve, and the median number of bids submitted in a curve is "),
      tags$strong(match_median),
      glue::glue(". "),
      tags$strong(match_single),
      glue::glue("% of {match_id}'s bids are single-point curves. "),
      tags$strong(match_self),
      glue::glue("% of this ID’s bids are self-scheduled. "),
      match_null_sentence,
      " ",
      match_negative_sentence
    )
  })

  div(
    class = "card",
    style = "padding: 22px; font-size: 18px; line-height: 1.7;",

    tags$h3(glue::glue("Bidding Behavior for ID {target_id}")),

    tags$p(
      glue::glue("ID {target_id} bidding behavior showed that the maximum MW quantity ever submitted in a bid was "),
      tags$strong(target_max_mw),
      " MW. The mean number of bids submitted in a curve is ",
      tags$strong(target_mean),
      " and the median number of bids submitted in a curve is ",
      tags$strong(target_median),
      ". ",
      tags$strong(target_single),
      glue::glue("% of {target_id}'s bids are single-point curves. "),
      tags$strong(target_self),
      "% of this ID’s bids are self-scheduled. ",
      target_null_sentence,
      " ",
      target_negative_sentence
    ),

    tags$hr(),

    tags$h3("Top Matching IDs"),

    match_paragraphs
  )
})
```

# NEW TEST PAGE

## Row
::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}

## ID to ID Matching

This page displays the most similar Resource IDs for the selected ID. The table on the left lists the top matching IDs ranked by similarity score. Similarity scores range from 0 to 1, with values closer to 1 indicating more similar bidding behavior. The table on the right displays the attributes used in the matching process, where you can compare the characteristics of the matched resources.

To view matches for a different resource, enter a new Resource ID on the **About** page and click **Submit ID**.
:::
## Row

### Column {width="30%"}

```{r}
uiOutput("target_id_summary")
```

```{r}
#| context: server

output$target_id_summary <- renderUI({

  req(input$resource_id)

  target_id <- as.character(input$resource_id)

  target_max_mw <- get_feature(features, target_id, "Max MW")
  target_mean <- get_feature(features, target_id, "Mean knots")
  target_median <- get_feature(features, target_id, "Median knots")
  target_single <- get_feature(features, target_id, "Single-point curves")
  target_self <- get_feature(features, target_id, "Self-scheduled bids")
  target_null <- get_feature(features, target_id, "Null bids")
  target_negative <- get_feature(features, target_id, "Negative bids")

  target_null_sentence <- has_text(target_null, target_id, "NULL bids")
  target_negative_sentence <- has_text(target_negative, target_id, "negative bids")

  div(
    class = "card",
    style = "
      padding: 24px;
      background-color: #FFFFFF;
        border: 6px solid #a2c7ea;
        border-radius: 8px;
    #   font-size: 17px;
    #   line-height: 1.7;
    ",

    tags$h4(
      style = "margin-top:0;",
      glue::glue("Bidding behavior for {target_id}:")
    ),

    tags$p(
      glue::glue(
        "The maximum MW quantity ever submitted in a bid was "
      ),
      tags$strong(style = "color:#18BC9C;", target_max_mw),
      " MW."
    ),

    tags$p(
      "The mean number of bids submitted in a curve is ",
      tags$strong(style = "color:#F39C12;", target_mean),
      " and the median number of bids submitted in a curve is ",
      tags$strong(style = "color:#EBBA5C;", target_median),
      "."
    ),

    tags$p(
      tags$strong(style = "color:#E74C3C;", target_single),
      glue::glue("% of ID {target_id}'s bids are single-point curves.")
    ),

    tags$p(
      tags$strong(style = "color:#E98A7F;", target_self),
      "% of this ID's bids are self-scheduled."
    ),

    tags$p(target_null_sentence),
    tags$p(target_negative_sentence)
  )
})
```

### Column {width="70%"}
```{r}
uiOutput("match_cards")
```

```{r}
#| context: server

output$match_cards <- renderUI({

  req(input$resource_id)

  best_matches_tbl <- best_matches()

  cards <- purrr::map(seq_len(nrow(best_matches_tbl)), function(i) {

    match_id <- best_matches_tbl$matched_id[i]
    sim <- round(best_matches_tbl$similarity[i], 3)

    match_max_mw <- get_feature(features, match_id, "Max MW")
    match_mean <- get_feature(features, match_id, "Mean knots")
    match_median <- get_feature(features, match_id, "Median knots")
    match_single <- get_feature(features, match_id, "Single-point curves")
    match_self <- get_feature(features, match_id, "Self-scheduled bids")
    match_null <- get_feature(features, match_id, "Null bids")
    match_negative <- get_feature(features, match_id, "Negative bids")

    match_null_sentence <- has_text(match_null, match_id, "NULL bids")
    match_negative_sentence <- has_text(match_negative, match_id, "negative bids")

    div(
    class = "card",
    style = "
        padding: 16px;
        margin-bottom: 12px;
        background-color: #ffffff;
        border-left: 6px solid #a2c7ea;
        # font-size: 15px;
        # line-height: 1.5;
    ",

    tags$strong(
        style = "font-size:16px; color=#18BC9C;",
        glue::glue("Match {i}: ID {match_id}")
    ),

    tags$p(
        style = "margin-top:8px; margin-bottom:0px;",

        "This ID has a similarity of ",
        tags$strong(style = "color:#3498DB;", sim),
        ". The maximum MW submitted in a bid is ",
        tags$strong(style = "color:#18BC9C;", match_max_mw),
        " MW. ID ",
        tags$strong(match_id),
        " submits on average ",
        tags$strong(style = "color:#F39C12;", match_mean),
        " bids in a curve, with the median number being ",
        tags$strong(style = "color:#EBBA5C;", match_median),
        "bids. ",
        tags$strong(style = "color:#E74C3C;", match_single),
        "% of its bids are single-point curves and ",
        tags$strong(style = "color:#E98A7F;", match_self),
        "% are self-scheduled. ",
        match_null_sentence,
        " ",
        match_negative_sentence
    )
    )
  })

  tagList(cards)
})
```

# ID FILTER

::: {.panel.tabset}

## Before

```{r}
uiOutput("before_match_cards")
```

## During

```{r}
uiOutput("during_match_cards")
```

## After

```{r}
uiOutput("after_match_cards")
```

```{r}
#| context: server

build_match_cards <- function(match_data) {

  cards <- purrr::map(seq_len(nrow(match_data)), function(i) {

    match_id <- match_data$matched_id[i]
    sim <- round(match_data$similarity[i], 3)

    start_date <- format(match_data$start[i], "%Y-%m-%d")
    end_date <- format(match_data$end[i], "%Y-%m-%d")

    match_max_mw <- get_feature(features, match_id, "Max MW")
    match_mean <- get_feature(features, match_id, "Mean knots")
    match_median <- get_feature(features, match_id, "Median knots")
    match_single <- get_feature(features, match_id, "Single-point curves")
    match_self <- get_feature(features, match_id, "Self-scheduled bids")

    match_null <- get_feature(features, match_id, "Null bids")
    match_negative <- get_feature(features, match_id, "Negative bids")

    match_null_sentence <- has_text(match_null, match_id, "NULL bids")
    match_negative_sentence <- has_text(match_negative, match_id, "negative bids")

    div(
      class = "card",
      style = "
        padding:16px;
        margin-bottom:12px;
        background-color:#FFFFFF;
        border-left:6px solid #a2c7ea;
      ",

      tags$strong(
        style = "font-size:16px;",
        glue::glue("ID {match_id}")
      ),

      tags$p(
        style = "margin-top:8px; margin-bottom:0px;",

        "This ID existed from ",
        tags$strong(start_date),
        " through ",
        tags$strong(end_date),
        ". It has a similarity of ",
        tags$strong(style = "color:#3498DB;", sim),
        ". The maximum MW submitted in a bid is ",
        tags$strong(style = "color:#18BC9C;", match_max_mw),
        " MW. ID ",
        tags$strong(match_id),
        " submits on average ",
        tags$strong(style = "color:#F39C12;", match_mean),
        " bids in a curve, with the median number being ",
        tags$strong(style = "color:#EBBA5C;", match_median),
        ". ",
        tags$strong(style = "color:#E74C3C;", match_single),
        "% of its bids are single-point curves and ",
        tags$strong(style = "color:#E98A7F;", match_self),
        "% are self-scheduled. ",
        match_null_sentence,
        " ",
        match_negative_sentence
      )
    )
  })

  tagList(cards)
}

output$before_match_cards <- renderUI({

  matches <- matches_by_period() |>
    filter(period == "Before")

  build_match_cards(matches)
})

output$during_match_cards <- renderUI({

  matches <- matches_by_period() |>
    filter(period == "During")

  build_match_cards(matches)
})

output$after_match_cards <- renderUI({

  matches <- matches_by_period() |>
    filter(period == "After")

  build_match_cards(matches)
})
```

# Average Bid Curve

##
::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## Average Bid Curve

Here you can view the average bid curve for market participants. This represents the typical daily bid behavior for a resource. Hover over plots to see more information.

To view the average bid curve for for a different resource, enter a new Resource ID on the **About** page and click **Submit ID**.
:::

##
### {.sidebar width="300px"}

```{r}
# selectInput(
#       inputId = "num_sd",
#       label = "Choose the number of standard deviations to determine outliers.",
#       choices = 
#         list(
#           "1 SD", 
#           "2 SD"
#         ),
#       selected = "2 SD"
#     )

dateRangeInput(
    inputId = "dates",
    label = "Date range to compute average:",
    start = "2023-01-01",
    end = "2025-06-07",
    separator = "to",
    min = "2023-01-01",
    max = "2025-06-07")

# radioButtons(
#     inputId = "num_sd",
#     label = "Choose the number of standard deviations to determine outliers.",
#     choices = list("1 SD" = 1, "2 SD" = 2),
#     selected = 2
#     )

sliderInput("num_sd", 
            label = "Choose the number of standard deviations to determine outliers.", min = 1, 
            max = 4, 
            value = 2)

checkboxGroupInput(
    inputId = "plot_display",
    label = "Choose the features to display on the plots.",
    choices = list("Standard deviation bands" = 1,
                   "Outlier bids" = 2),
    selected = c(1,2)
)

# checkboxGroupInput(
#   inputId = "hours_display",
#   label = "Choose the hours to display on the plots.",
#   choices = setNames(
#     0:23,
#     sprintf("%02d:00", 0:23)
#   ),
#   selected = 16:20
# )

selectizeInput(
  inputId = "hours_display",
  label = "Choose hours:",
  choices = setNames(0:23, sprintf("%02d:00", 0:23)),
  selected = 16:20,
  multiple = TRUE
)
```

```{r}
#| context: server

output$id_warning <- renderUI({

  req(input$resource_id)

  # valid_ids <- unique(interp_tbl$RESOURCEBID_SEQ |>)
  valid_ids <- interp_tbl |>
    select(RESOURCEBID_SEQ) |>
    distinct() |>
    collect() |>
    pull(RESOURCEBID_SEQ)

  if (!(input$resource_id %in% valid_ids)) {

    div(
      style = "color: red; font-weight: bold;",
      "Please enter a valid 6-digit ID that exists in the DAM during 1/2023 - 6/2025."
    )

  }

})
```

###

::: {.panel-tabset}

#### Overall Daily Trend

```{r}
plotlyOutput("overall_plot")
```

```{r}
#| context: server

output$overall_plot <- renderPlotly({

    num_sd <- as.numeric(input$num_sd)
    avg_curve <- average_bid_curve()

    rects <- data.frame(
        xmin = seq(0, 2300, by = 200),
        xmax = seq(100, 2400, by = 200),
        ymin = -Inf,
        ymax = Inf
    )
    overall_plot <- ggplot() +
      geom_rect(data = rects,
                aes(xmin = xmin, xmax = xmax, ymin=ymin, ymax = ymax),
                  fill = "grey50",
                    alpha = 0.1,
                    inherit.aes = FALSE) 
    
    if (1 %in% input$plot_display){
      overall_plot <- overall_plot +
        geom_ribbon(data = avg_curve, 
                    aes(x = daily_x,
                        ymin = avg_price - sd*num_sd,
                        ymax = avg_price + sd*num_sd),
                    fill = "#3498DB",
                    alpha = 0.2)
    }

    overall_plot <- overall_plot +
      geom_step(data = avg_curve, 
                aes(x = daily_x, 
                    y = avg_price,
                    group = 1,
                    text = glue::glue("Hour: {hour}<br>",
                                      "Quantity: {orig_MW} MW<br>",
                                      "Price: ${round(avg_price, 2)}"))) +
      theme_bw() +
      labs(
        x = "Hour",
        y = "Price ($/MWh)",
        title = glue::glue("Average Daily Bid Curve for {input$resource_id}")
        # subtitle = glue::glue("Resource ID: {input$resource_id}")
      ) +
      scale_x_continuous(breaks = seq(0, 2300, by = 100),
                      labels = seq(0, 23, by = 1),
                  expand = c(0, 0)) +
      theme(panel.grid.minor.x = element_blank(),
            # plot.title = element_text(size = 20),
            # axis.text = element_text(size = 16),
            # axis.title = element_text(size = 18)
            )   

    ggplotly(overall_plot, tooltip = "text") 
    # |>
    #   layout(
    #     title = list(
    #     text = glue::glue(
    #       "Average Daily Bid Curve<br>",
    #       "<sup>Resource ID: {input$resource_id} from {input$dates[1]} to {input$dates[2]}</sup>"
    #     )),
    #     x=0
    #   )
})

```

<!-- ::: {.callout-note appearance="minimal"}

Note abt the graph

::: -->

#### Hourly Curves

::: {.panel-tabset}

##### Combined

```{r}
plotlyOutput("hourly_plot")
```

```{r}
#| context: server
#| height: "100%"
#| fill: true

n_colors <- 24
gradient_colors <- c("#85d9b1", "#357ba2ff", "#2B3E50")
discrete_colors <- setNames(colorRampPalette(gradient_colors)(n_colors), 0:23)

output$hourly_plot <- renderPlotly({

  avg_curve <- average_bid_curve() |>
    filter(hour %in% as.numeric(input$hours_display)
    )

  hourly_plot <- avg_curve |>
    ggplot(aes(x = orig_MW, y = avg_price, group = hour, color = as.factor(hour))) +
    geom_step(aes(text = glue::glue("Hour: {hour}<br>",
                                    "Quantity: {orig_MW} MW<br>",
                                    "Price: ${round(avg_price, 2)}")),
                                    linewidth = 0.5) +
    # geom_point(aes(text = glue::glue("Hour: {hour}<br>",
    #                                  "Quantity: {pct_grid}%<br>",
    #                                  "Price: ${round(avg_price, 2)}"))) +
    theme_bw() +
    labs(x = "Quantity (MW)",
        y = "Price ($/MWh)",
        title = glue::glue("Average Bid Curve by Hour for {input$resource_id}"),
        color = "Hour") +
    # scale_x_continuous(breaks = seq(0, 100, by = 10)) +
    # scale_color_gradient2(low = "#85d9b1",
    #                       mid = "#357ba2ff",
    #                       high = "#2B3E50",
    #                       midpoint = 12) +
    scale_color_manual(values = discrete_colors) +
    # scale_color_viridis_c(option = "mako",
    #                       direction = -1) +
    theme(panel.grid.minor.x = element_blank())

  ggplotly(hourly_plot, tooltip = "text")
})
```

##### Separated

```{r}
plotlyOutput("faceted_p")
```
```{r}
#| context: server
hour_labels <- c(
       `0` = "00:00",
       `1` = "1:00",
       `2` = "2:00",
       `3` = "3:00",
       `4` = "4:00",
       `5` = "5:00",
       `6` = "6:00",
       `7` = "7:00",
       `8` = "8:00",
       `9` = "9:00",
       `10` = "10:00",
       `11` = "11:00",
       `12` = "12:00",
       `13` = "13:00",
       `14` = "14:00",
       `15` = "15:00",
       `16` = "16:00",
       `17` = "17:00",
       `18` = "18:00",
       `19` = "19:00",
       `20` = "20:00",
       `21` = "21:00",
       `22` = "22:00",
       `23` = "23:00"
)

output$faceted_p <- renderPlotly({

  num_sd <- as.numeric(input$num_sd)
  avg_curve <- average_bid_curve()

  faceted_plot <- avg_curve |>
    ggplot(aes(x = orig_MW, y = avg_price, group = hour)) +
    geom_step(aes(text = glue::glue("Price: ${round(avg_price)}<br>",
                                    "Quantity: {orig_MW} MW"))) +
    theme_bw() +
    labs(
      x = "Quantity (MW)",
      y = "Price ($/MWh)",
      title = glue::glue("Average Daily Bid Curve for {input$resource_id}")
    ) +
    facet_wrap(~hour, 
              nrow = 3,
              labeller = as_labeller(hour_labels)) +
    # scale_x_continuous(breaks = seq(0, 100, by = 20)) +
    theme(panel.grid.minor.x = element_blank())

  if (1 %in% input$plot_display){
      faceted_plot <- faceted_plot +
        geom_ribbon(data = avg_curve, 
                    aes(x = orig_MW,
                        ymin = if_else(avg_price - num_sd*sd < 0, 0, avg_price - num_sd*sd),
                        ymax = avg_price + 2*sd),
                    fill = "#3498DB",
                    alpha = 0.2)
    }

  ggplotly(faceted_plot, tooltip = "text") |> config(responsive = TRUE)
})

```

<!-- ::: {.callout-note appearance="minimal"}

Note abt the graph

::: -->

:::

:::