---
title: "Bid Curve Shenanigans"
subtitle: ""
format: 
  dashboard:
    orientation: rows
    theme: flatly
server: shiny
embed-resources: true
---

<style>
.navbar,
.quarto-dashboard-header,
header {
  background-color: #c42e1d !important;
}
</style>

```{r}
#| label: load-packages
#| include: false
#| context: setup
library(tidyverse)
library(plotly)
library(duckdb)
library(gt)
library(Matrix)
```

```{r}
#| label: load-data
#| include: false
#| context: setup
con <- dbConnect(duckdb())
tbl_pointer <- tbl(con, "read_parquet('CAISO_rtm_big.parquet')")
interp_tbl <- arrow::open_dataset("interp_dataset_rtm")

gen_matches_all <- read_csv("final_outputs/outage_rtm_top_candidates.csv") |>
  mutate(
    `RESOURCE ID` = generator,
    `RESOURCE NAME` = generator,
    `RESOURCE PMAX MW` = pmax_mw,
    seq_max_mw = max_mw,
    confidence = NA_real_
  )

best_gen_match <- read_csv("final_outputs/outage_rtm_matches_final.csv") |>
  mutate(
    `RESOURCE ID` = generator,
    `RESOURCE NAME` = generator,
    `RESOURCE PMAX MW` = pmax_mw,
    seq_max_mw = max_mw
  )

outage_events <- read_csv("final_outputs/outage_rtm_event_counts.csv")
```

```{r}
#| context: server
max_mw <- eventReactive(input$submit_id, {
  req(input$resource_id)

  tbl_pointer |>
    filter(RESOURCEBID_SEQ == input$resource_id) |>
    summarise(max_mw = max(SCH_BID_XAXISDATA, na.rm = TRUE)) |>
    collect() |>
    pull(max_mw)
})


average_bid_curve <- eventReactive(input$submit_id, {
  req(input$resource_id)

  max_mw2 <- max_mw()

  interp_tbl |>
    filter(
      RESOURCEBID_SEQ == input$resource_id,
      date(date) >= input$dates[1],
      date(date) <= input$dates[2]
    ) |>
    # group by each hour each day
    group_by(hour, pct_grid, daily_x) |>
    summarise(
      avg_price = mean(price_interp, na.rm = TRUE),
      sd = sd(price_interp, na.rm = TRUE),
      # number of days included in the average
      n_days_used = sum(!is.na(price_interp)),
      .groups = "drop"
    ) |>
    # if no days included in avg. filter out
    filter(n_days_used != 0) |>
    collect() |>
    mutate(orig_MW = (pct_grid / 100) * max_mw2)
})


gen_match_candidates <- eventReactive(input$submit_id, {
  req(input$resource_id)

  target_id <- as.numeric(input$resource_id)

  best_match <- best_gen_match |>
    filter(RESOURCEBID_SEQ == input$resource_id)

  exclude_id <- best_match$`RESOURCE ID`

  other_matches <- gen_matches_all |>
    filter(
      RESOURCEBID_SEQ == input$resource_id,
      !(`RESOURCE ID` %in% exclude_id)
    ) |>
    arrange(desc(!is.na(confidence)), desc(score)) |>
    slice_head(n = 4)

  bind_rows(
    best_match,
    other_matches
  ) |>
    left_join(
      outage_events |>
        select(`RESOURCE ID`, total_outage_events),
      by = "RESOURCE ID"
    )
})

output$id_warning <- renderUI({
  req(input$resource_id)

  valid_ids <- c(
    best_gen_match$RESOURCEBID_SEQ,
    gen_matches_all$RESOURCEBID_SEQ
  ) |>
    unique()

  if (!(input$resource_id %in% valid_ids)) {
    div(
      style = "color: red; font-weight: bold;",
      "Please enter a valid RTM ID with outage matching results."
    )
  }
})
```

# About

::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## CAISO RTM Outage Matching

Here you can view average bid curves and outage-based deanonymization results for market participants in CAISO. Enter an ID to view its average bid curve in the **Average Bid Curve** tab and potential resource-name matches for January 2023 to June 2025 in the **Deanonymization** tab. Hover over words highlighted in <span style="color:#325D88;">blue</span> for more information.
:::


## Row
::: {.card width="20%"}
## Explore a Resource ID

To populate the Average Bid Curve and Deanonymization with results, first submit the anonymized RTM ID you want to view results for.

```{r}
numericInput(
  inputId = "resource_id",
  label = "Please enter a RTM Resource ID:",
  value = NULL
)

actionButton(
  inputId = "submit_id",
  label = "Submit ID",
  class = "btn-primary"
)

uiOutput("id_warning")
```

:::

::: {.card}

## **Navigate the dashboard**

::: {.panel-tabset}

# Average Bid Curve

To compare bid curves across days, each ID's MW values are rescaled to a common percentage scale and averaged across daily bidding patterns. We average RTM bids to hour resolution.

1. **Rescale MW values**  
   Convert each submitted MW quantity into a percentage of the resource’s maximum submitted MW.

2. **Round to a 1% grid**  
   Round each MW percentage to the nearest whole percent.

3. **Interpolate within each hourly curve**  
   For each day-hour curve, fill in prices on a 0–100% MW grid.

   - Use the actual submitted price when one exists.
   - Otherwise, use the closest lower submitted MW percentage.
   - Do not extrapolate beyond the observed MW range for that hour.

4. **Stack hourly curves into a daily curve**  
   Combine the 24 hourly curves into one continuous daily curve.

5. **Average across days**  
   Average prices across all days at each hour-percent combination to create the resource’s typical 24-hour bid curve.

# Deanonymization

When a generator has a reported full outage, its RTM bid submissions should disappear for the outage window. This pipeline matches reported outage windows to gaps in RTM bid submissions for anonymous IDs. Further filtering is based on the number of matched outage events, the closeness of the ID's max bid MW to the resource's pmax, and timing offsets around outage stop and restart. A score and a confidence are given for the matches. The score is based on the mean absolute percentage error of bids across all matching days along with the number of matches. The confidence (0-1 scale) is based on how much the top score is better than the runner up. A score of 1 represents an unambiguous match.

:::

:::

# Average Bid Curve

##
::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## Average Bid Curve

Here you can view the average bid curve for market participants. This represents the typical daily bid behavior for a resource. Hover over plots to see more information.

To view the average bid curve for for a different resource, enter a new Resource ID on the **About** page and click **Submit ID**.
:::

##
### {.sidebar width="300px"}

```{r}
dateRangeInput(
  inputId = "dates",
  label = "Date range to compute average:",
  start = "2023-01-01",
  end = "2025-06-07",
  separator = "to",
  min = "2023-01-01",
  max = "2025-06-07"
)

sliderInput(
  "num_sd",
  label = "Choose the number of standard deviations to determine outliers.",
  min = 1,
  max = 4,
  value = 2
)

checkboxGroupInput(
  inputId = "plot_display",
  label = "Choose the features to display on the plots.",
  choices = list("Standard deviation bands" = 1, "Outlier bids" = 2),
  selected = c(1, 2)
)

selectizeInput(
  inputId = "hours_display",
  label = "Choose hours:",
  choices = setNames(0:23, sprintf("%02d:00", 0:23)),
  selected = 16:20,
  multiple = TRUE
)
```

###

::: {.panel-tabset}

#### Overall Daily Trend

```{r}
plotlyOutput("overall_plot")
```

```{r}
#| context: server
output$overall_plot <- renderPlotly({
  num_sd <- as.numeric(input$num_sd)
  avg_curve <- average_bid_curve()

  rects <- data.frame(
    xmin = seq(0, 2300, by = 200),
    xmax = seq(100, 2400, by = 200),
    ymin = -Inf,
    ymax = Inf
  )
  overall_plot <- ggplot() +
    geom_rect(
      data = rects,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      fill = "grey50",
      alpha = 0.1,
      inherit.aes = FALSE
    )

  if (1 %in% input$plot_display) {
    overall_plot <- overall_plot +
      geom_ribbon(
        data = avg_curve,
        aes(
          x = daily_x,
          ymin = avg_price - sd * num_sd,
          ymax = avg_price + sd * num_sd
        ),
        fill = "#3498DB",
        alpha = 0.2
      )
  }

  overall_plot <- overall_plot +
    geom_step(
      data = avg_curve,
      aes(
        x = daily_x,
        y = avg_price,
        group = 1,
        text = glue::glue(
          "Hour: {hour}<br>",
          "Quantity: {orig_MW} MW<br>",
          "Price: ${round(avg_price, 2)}"
        )
      )
    ) +
    theme_bw() +
    labs(
      x = "Hour",
      y = "Price ($/MWh)",
      title = glue::glue("Average Daily Bid Curve for {input$resource_id}")
    ) +
    scale_x_continuous(
      breaks = seq(0, 2300, by = 100),
      labels = seq(0, 23, by = 1),
      expand = c(0, 0)
    ) +
    theme(panel.grid.minor.x = element_blank())

  ggplotly(overall_plot, tooltip = "text")
})
```

<!-- ::: {.callout-note appearance="minimal"}

Note abt the graph

::: -->

#### Hourly Curves

::: {.panel-tabset}

##### Combined

```{r}
plotlyOutput("hourly_plot")
```

```{r}
#| context: server
#| height: "100%"
#| fill: true
n_colors <- 24
gradient_colors <- c("#85d9b1", "#357ba2ff", "#2B3E50")
discrete_colors <- setNames(colorRampPalette(gradient_colors)(n_colors), 0:23)

output$hourly_plot <- renderPlotly({
  avg_curve <- average_bid_curve() |>
    filter(hour %in% as.numeric(input$hours_display))

  hourly_plot <- avg_curve |>
    ggplot(aes(
      x = orig_MW,
      y = avg_price,
      group = hour,
      color = as.factor(hour)
    )) +
    geom_step(
      aes(
        text = glue::glue(
          "Hour: {hour}<br>",
          "Quantity: {orig_MW} MW<br>",
          "Price: ${round(avg_price, 2)}"
        )
      ),
      linewidth = 0.5
    ) +
    theme_bw() +
    labs(
      x = "Quantity (MW)",
      y = "Price ($/MWh)",
      title = glue::glue("Average Bid Curve by Hour for {input$resource_id}"),
      color = "Hour"
    ) +
    scale_color_manual(values = discrete_colors) +
    theme(panel.grid.minor.x = element_blank())

  ggplotly(hourly_plot, tooltip = "text")
})
```

##### Separated

```{r}
plotlyOutput("faceted_p")
```
```{r}
#| context: server
hour_labels <- c(
  `0` = "00:00",
  `1` = "1:00",
  `2` = "2:00",
  `3` = "3:00",
  `4` = "4:00",
  `5` = "5:00",
  `6` = "6:00",
  `7` = "7:00",
  `8` = "8:00",
  `9` = "9:00",
  `10` = "10:00",
  `11` = "11:00",
  `12` = "12:00",
  `13` = "13:00",
  `14` = "14:00",
  `15` = "15:00",
  `16` = "16:00",
  `17` = "17:00",
  `18` = "18:00",
  `19` = "19:00",
  `20` = "20:00",
  `21` = "21:00",
  `22` = "22:00",
  `23` = "23:00"
)

output$faceted_p <- renderPlotly({
  num_sd <- as.numeric(input$num_sd)
  avg_curve <- average_bid_curve()

  faceted_plot <- avg_curve |>
    ggplot(aes(x = orig_MW, y = avg_price, group = hour)) +
    geom_step(aes(
      text = glue::glue(
        "Price: ${round(avg_price)}<br>",
        "Quantity: {orig_MW} MW"
      )
    )) +
    theme_bw() +
    labs(
      x = "Quantity (MW)",
      y = "Price ($/MWh)",
      title = glue::glue("Average Daily Bid Curve for {input$resource_id}")
    ) +
    facet_wrap(~hour, nrow = 3, labeller = as_labeller(hour_labels)) +
    theme(panel.grid.minor.x = element_blank())

  if (1 %in% input$plot_display) {
    faceted_plot <- faceted_plot +
      geom_ribbon(
        data = avg_curve,
        aes(
          x = orig_MW,
          ymin = if_else(
            avg_price - num_sd * sd < 0,
            0,
            avg_price - num_sd * sd
          ),
          ymax = avg_price + 2 * sd
        ),
        fill = "#3498DB",
        alpha = 0.2
      )
  }

  ggplotly(faceted_plot, tooltip = "text") |> config(responsive = TRUE)
})
```

<!-- ::: {.callout-note appearance="minimal"}

Note abt the graph

::: -->

:::

:::
# Deanonymization

::: {.callout-note appearance="minimal" style="background-color: #F8F8F8; border-left: 6px solid #2B3E50;"}
## ID to Resource Name

This page displays the most similar resources for the selected ID using outage matching. Matches are ordered in terms of confidence. Hover over words highlighted in <span style="color:#325D88;">blue</span> for more information.

To view matches for a different ID, enter a new Resource ID on the **About** page and click **Submit ID**.
:::
<!-- 
he table on the left lists definitions of the attributes using in matching. The table on the right displays the matched resource along with matching scores and information about the resource used in the matching process. -->

## Row

<!-- ### Column  -->

<!-- ::: {.card}

## Matching Attribute Definitions

- **Resource Name:** Official name of the matched resource.
- **Resource Pmax:** CAISO reported Physical Maximum output for the matched resource.
- **ID Max MW:** Anonymized ID’s all time maximum MW bid
- **# of Matches:** Number of matched outage events.
- **Score:** Number of Matches x (1 - mean absolute percentage error across all matching days)
- **Confidence:** 0-1 scale of how much better the top score compared to the runner up (1 being an unambiguous match)

::: -->

### Column

```{r}
uiOutput("gen_match_summary")
```

```{r}
#| context: server
output$gen_match_summary <- renderUI({
  req(input$resource_id)

  target_id <- as.character(input$resource_id) |> str_trim()

  gen_rows <- gen_match_candidates()

  validate(
    need(nrow(gen_rows) > 0, "No deanonymization matches found for this ID.")
  )

  rank_labels <- c(
    "most likely",
    "second most likely",
    "third most likely",
    "fourth most likely",
    "fifth most likely"
  )
  cards <- purrr::map(seq_len(nrow(gen_rows)), function(i) {
    row <- gen_rows[i, ]

    resource_name <- row$`RESOURCE NAME`
    resource_pmax <- round(row$`RESOURCE PMAX MW`, 3)
    id_max_mw <- round(row$seq_max_mw, 3)
    n_matches <- row$n_matches
    total_events <- row$total_outage_events
    score <- round(row$score, 3)
    confidence <- round(row$confidence, 3)

    confidence_text <- if (!is.na(confidence)) {
      tagList(
        ", with a ",
        tags$span(
          "confidence",
          title = "How much better the top score is compared to the runner-up. Values closer to 1 indicate a clearer match. Confidence is only used between the top two matches.",
          style = "color:#325D88; cursor:help;"
        ),
        " of ",
        tags$strong(confidence, style = "color:#3498DB;")
      )
    } else {
      NULL
    }

    div(
      class = "card",
      style = "
        padding: 18px;
        margin-bottom: 14px;
        background-color: #FFFFFF;
        border-left: 6px solid #a2c7ea;
        font-size: 16px;
        line-height: 1.6;
      ",

      p(
        style = "margin-top:8px; margin-bottom:0;",

        "The ",
        rank_labels[i],
        " ",
        tags$span(
          "resource",
          title = "Official name of the matched resource.",
          style = "color:#325D88; cursor:help;"
        ),
        " for anonymous ID ",
        tags$strong(target_id),
        " is ",
        tags$strong(resource_name, style = "color:#F39C12;"),
        ". Our final ",
        tags$span(
          "score",
          title = "Number of matches multiplied by (1 - the mean absolute percentage error).",
          style = "color:#325D88; cursor:help;"
        ),
        " for this match is ",
        tags$strong(score, style = "color:#3498DB;"),
        confidence_text,
        ". Why? The resource ",
        tags$span(
          "Pmax",
          title = "CAISO reported physical maximum output for the matched resource.",
          style = "color:#325D88; cursor:help;"
        ),
        " is ",
        tags$strong(paste0(resource_pmax, " MW"), style = "color:#18BC9C;"),
        ", and ID ",
        target_id,
        "'s maximum MW ever submitted in a bid is ",
        tags$strong(paste0(id_max_mw, " MW"), style = "color:#18BC9C;"),
        ". Out of ",
        tags$strong(style = "color:#E74C3C;", total_events),
        " total outage events, ",
        tags$strong(style = "color:#E98A7F;", n_matches),
        " matched this anonymous ID."
      )
    )
  })

  tagList(cards)
})
```
