---
title: "Resource ID Feature"
format: html
---

```{r}
library(duckdb)
library(tidyverse)
```


```{r}
con <- dbConnect(duckdb())
tbl_pointer <- tbl(con, "read_parquet('CAISO_dam_big.parquet')")
```

```{r}
data <- tbl_pointer |>
    filter(MARKETPRODUCTTYPE == "EN",
           RESOURCE_TYPE == "GENERATOR",
           !is.na(RESOURCEBID_SEQ)) |>
    select(RESOURCEBID_SEQ, STARTTIME, SCH_BID_XAXISDATA, SCH_BID_Y1AXISDATA, SELFSCHEDMW) |>
    collect()
```

Get max MW
```{r}
observed_max <- data |>
    group_by(RESOURCEBID_SEQ) |>
    # filter(!is.na(SCH_BID_XAXISDATA)) |>
    slice_max(n=1, order_by = SCH_BID_XAXISDATA, with_ties = FALSE) |>
    select(RESOURCEBID_SEQ, SCH_BID_XAXISDATA)
```

Proportion of bid curves with negative MW

```{r}
negative_mw <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    summarise(has_negative_mw = any(SCH_BID_XAXISDATA < 0, na.rm = TRUE), .groups = "drop") |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(pct_negative_mw = 100 * mean(has_negative_mw), .groups = "drop")
```

Mean and median number of knots
```{r}
knots <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    summarise(knots = n()) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(mean = mean(knots),
              median = median(knots))
```

single-point frequency

```{r}
single <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    mutate(knots = n(),
           single_bid_hour = knots == 1) |>
    ungroup() |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(total_bids = n(),
              num_single = sum(single_bid_hour),
              pct_single = 100*num_single/total_bids) |>
    select(RESOURCEBID_SEQ, pct_single)
```

% of all bids that are NULL bids

```{r}
null <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    summarise(
        null_bid = any(is.na(SCH_BID_XAXISDATA) & is.na(SCH_BID_Y1AXISDATA)),
        .groups = "drop"
    ) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(null = 100 * mean(null_bid), .groups = "drop")
```

% of self scheduled bids

```{r}
self <- data |>
    mutate(self_sched = !is.na(SELFSCHEDMW)) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(total_bids = n(),
              pct_self_sched = 100 * mean(self_sched)) |>
    select(RESOURCEBID_SEQ, pct_self_sched)
```

hours_tbl

```{r}
hours_tbl <- data |>
    mutate(hour = formatC(hour(ymd_hms(STARTTIME)), width = 2, flag = "0")) |>
    select(RESOURCEBID_SEQ, hour) |>
    group_by(RESOURCEBID_SEQ, hour) |>
    summarise(num_bids = n(), .groups = "drop") |>
    group_by(RESOURCEBID_SEQ) |>
    mutate(pct_per_hour = 100*num_bids / sum(num_bids)) |>
    select(!num_bids) |>
    pivot_wider(names_from = hour,
                values_from = pct_per_hour,
                names_prefix = "hour_",
                names_sort = TRUE,
                values_fill = 0)
```

dates of start and end

```{r}
start_end <- data |>
    select(RESOURCEBID_SEQ, STARTTIME) |>
    mutate(STARTTIME = as.Date(STARTTIME)) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(start = min(STARTTIME),
              end = max(STARTTIME), .groups = "drop")
```

join everything

```{r}
features <- knots |>
    left_join(single, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(null, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(self, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(start_end, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(negative_mw, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(observed_max, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |> 
    left_join(hours_tbl, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    arrow::write_parquet("ID_attributes.parquet")
```


filter MARKETPRODUCTTYPE = EN and resource_type
median # 

date start end 

hour - what % of all bids did they bid during each hour (24 features)

what % of bids are single point bids

null - yes/no

% of self scheduled bids


Simple similarity metrics for IDS.
Normalize features and calculate cosine similarity.

```{r}
# numeric matrix of features
feat_mat <- features |>
    mutate(
        days_active = as.numeric(end - start),
        max_mw = replace_na(SCH_BID_XAXISDATA, 0)
    ) |>
    select(-RESOURCEBID_SEQ, -start, -end, -SCH_BID_XAXISDATA) |>
    as.matrix()

# Scale each column to zero-mean, unit-variance
feat_scaled <- scale(feat_mat)
# Replace any NaN columns (zero-variance) with 0
feat_scaled[is.nan(feat_scaled)] <- 0

# Normalize each row to unit length
row_norms <- sqrt(rowSums(feat_scaled^2))
row_norms[row_norms == 0] <- 1
feat_norm <- feat_scaled / row_norms

# Cosine similarity = X %*% t(X) on row-normalized data
cos_sim <- tcrossprod(feat_norm)

ids <- features$RESOURCEBID_SEQ
rownames(cos_sim) <- ids
colnames(cos_sim) <- ids
cos_sim["100031", "177696"]
```

```{r}
find_matches <- function(ID, num_matches, similarities){
    tmp <- similarities[ID |> as.character(), ]
    tmp[order(abs(tmp), decreasing = TRUE)] |> head(num_matches + 1) |> tail(-1)
}

find_matches(653256, 5, cos_sim)
```

```{r}
lookup_id_features <- function(ID, features_tbl = features) {
    idx <- match(ID, features_tbl$RESOURCEBID_SEQ)

    if (is.na(idx)) {
        stop(glue::glue("RESOURCEBID_SEQ {ID} was not found in features_tbl."))
    }

    row <- features_tbl[idx, , drop = FALSE]
    fmt <- \(x) sprintf("%.2f", x)

    glue::glue(
        "ID: {ID}\n",
        "max MW: {fmt(row$SCH_BID_XAXISDATA[[1]])}\n",
        "mean # of bid pairs: {fmt(row$mean[[1]])}\n",
        "median # of bid pairs: {fmt(row$median[[1]])}\n",
        "single-point frequency: {fmt(row$pct_single[[1]])}\n",
        "NULL behavior: {fmt(row$null[[1]])}\n",
        "self-schedule behavior: {fmt(row$pct_self_sched[[1]])}\n",
        "negative bid behavior: {fmt(row$pct_negative_mw[[1]])}"
    )
}
lookup_id_features(653256)

hourly_plot <- function(ID, features_tbl = features) {
    features_tbl |>
        filter(RESOURCEBID_SEQ == ID) |>
        select(hour_00:hour_23) |>
        pivot_longer(cols = everything(), names_to = "hour", values_to = "pct_bids") |>
        mutate(
            hour = str_remove(hour, "^hour_"),
            hour = factor(hour, levels = sprintf("%02d", 0:23))
        ) |>
        ggplot(aes(x = hour, y = pct_bids)) +
        geom_col(fill = "#2C6E49") +
        scale_y_continuous(labels = scales::label_percent(scale = 1)) +
        labs(
            title = glue::glue("Hourly Bidding Pattern for ID {ID}"),
            x = NULL,
            y = NULL
        ) + 
        theme_bw()
}
hourly_plot(653256)
hourly_plot(985335)
hourly_plot(435045)
```
