---
title: "Resource ID Feature"
format: html
---

```{r}
library(duckdb)
library(tidyverse)
```


```{r}
con <- dbConnect(duckdb())
tbl_pointer <- tbl(con, "read_parquet('CAISO_dam_big.parquet')")
```

```{r}
data <- tbl_pointer |>
    filter(MARKETPRODUCTTYPE == "EN",
           RESOURCE_TYPE == "GENERATOR",
           !is.na(RESOURCEBID_SEQ)) |>
    select(RESOURCEBID_SEQ, STARTTIME, SCH_BID_XAXISDATA, SCH_BID_Y1AXISDATA, SELFSCHEDMW) |>
    collect()
```

Get max MW
```{r}
observed_max <- data |>
    group_by(RESOURCEBID_SEQ) |>
    # filter(!is.na(SCH_BID_XAXISDATA)) |>
    slice_max(n=1, order_by = SCH_BID_XAXISDATA, with_ties = FALSE) |>
    select(RESOURCEBID_SEQ, SCH_BID_XAXISDATA)
```

Mean and median number of knots
```{r}
knots <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    summarise(knots = n()) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(mean = mean(knots),
              median = median(knots))
```

single-point frequency

```{r}
single <- data |>
    group_by(STARTTIME, RESOURCEBID_SEQ) |>
    mutate(knots = n(),
           single_bid_hour = knots == 1) |>
    ungroup() |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(total_bids = n(),
              num_single = sum(single_bid_hour),
              pct_single = 100*num_single/total_bids) |>
    select(RESOURCEBID_SEQ, pct_single)
```

% of all bids that are NULL bids

```{r}
null <- data |>
    filter(is.na(SELFSCHEDMW)) |>
    select(RESOURCEBID_SEQ, SCH_BID_XAXISDATA, SCH_BID_Y1AXISDATA, SELFSCHEDMW) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(null = any((is.na(SCH_BID_XAXISDATA)) & (is.na(SCH_BID_Y1AXISDATA))))
```

% of self scheduled bids

```{r}
self <- data |>
    mutate(self_sched = !is.na(SELFSCHEDMW)) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(total_bids = n(),
              self_sched = sum(self_sched),
              pct_self_sched = 100 * mean(self_sched)) |>
    select(RESOURCEBID_SEQ, pct_self_sched)
```

hours_tbl

```{r}
hours_tbl <- data |>
    mutate(hour = hour(ymd_hms(STARTTIME))) |>
    select(RESOURCEBID_SEQ, hour) |>
    group_by(RESOURCEBID_SEQ, hour) |>
    summarise(num_bids = n(), .groups = "drop") |>
    group_by(RESOURCEBID_SEQ) |>
    mutate(pct_per_hour = 100*num_bids / sum(num_bids)) |>
    select(!num_bids) |>
    pivot_wider(names_from = hour,
                values_from = pct_per_hour,
                names_prefix = "hour_",
                values_fill = 0)

```

dates of start and end

```{r}
start_end <- data |>
    select(RESOURCEBID_SEQ, STARTTIME) |>
    mutate(STARTTIME = as.Date(STARTTIME)) |>
    group_by(RESOURCEBID_SEQ) |>
    summarise(start = min(STARTTIME),
              end = max(STARTTIME), .groups = "drop")
```

join everything

```{r}
features <- knots |>
    left_join(single, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(null, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(self, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(hours_tbl, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(start_end, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ)) |>
    left_join(observed_max, by = join_by(RESOURCEBID_SEQ == RESOURCEBID_SEQ))
```

```{r}
hours_tbl
```

filter MARKETPRODUCTTYPE = EN and resource_type
median # 

date start end 

hour - what % of all bids did they bid during each hour (24 features)

what % of bids are single point bids

null - yes/no

% of self scheduled bids


Simple similarity metrics for IDS.
Normalize features and calculate cosine similarity.

```{r}
# numeric matrix of features
feat_mat <- features |>
    mutate(
        null = replace_na(as.numeric(null), 0),
        days_active = as.numeric(end - start),
        max_mw = replace_na(SCH_BID_XAXISDATA, 0)
    ) |>
    select(-RESOURCEBID_SEQ, -start, -end, -SCH_BID_XAXISDATA) |>
    as.matrix()

# Scale each column to zero-mean, unit-variance
feat_scaled <- scale(feat_mat)
# Replace any NaN columns (zero-variance) with 0
feat_scaled[is.nan(feat_scaled)] <- 0

# Normalize each row to unit length
row_norms <- sqrt(rowSums(feat_scaled^2))
row_norms[row_norms == 0] <- 1
feat_norm <- feat_scaled / row_norms

# Cosine similarity = X %*% t(X) on row-normalized data
cos_sim <- tcrossprod(feat_norm)

ids <- features$RESOURCEBID_SEQ
rownames(cos_sim) <- ids
colnames(cos_sim) <- ids

cos_sim
```


```{r}
cos_sim["100031", "177696"]
```

Euclidean distance on z-scored features.

```{r}
compute_euclidean <- function(features_df) {
    mat <- features_df |>
        mutate(
            null = replace_na(as.numeric(null), 0),
            days_active = as.numeric(end - start),
            max_mw = replace_na(SCH_BID_XAXISDATA, 0)
        ) |>
        select(-RESOURCEBID_SEQ, -start, -end, -SCH_BID_XAXISDATA) |>
        as.matrix()

    # Scale each column to zero mean
    scaled <- scale(mat)
    scaled[is.nan(scaled)] <- 0

    dist_mat <- as.matrix(dist(scaled, method = "euclidean"))

    ids <- features_df$RESOURCEBID_SEQ
    rownames(dist_mat) <- ids
    colnames(dist_mat) <- ids

    dist_mat
}

euc_dist <- compute_euclidean(features)
euc_dist
```

```{r}
euc_dist["100031", "177696"]
```

```{r}
# Find the most similar generators to a given resource ID.
find_similar <- function(id, metric = c("euclidean", "cosine"), n = 10) {
    metric <- match.arg(metric)
    id <- as.character(id)

    mat <- switch(metric,
        euclidean = euc_dist,
        cosine = cos_sim
    )

    stopifnot(id %in% rownames(mat))

    scores <- mat[id, ]
    scores <- scores[names(scores) != id]

    if (metric == "euclidean") {
        sorted <- sort(scores, decreasing = FALSE)
    } else {
        sorted <- sort(scores, decreasing = TRUE)
    }

    tibble(
        RESOURCEBID_SEQ = names(sorted),
        score           = unname(sorted)
    ) |>
        slice_head(n = n)
}

find_similar("100031", metric = "euclidean")
find_similar("100031", metric = "cosine")
```