---
title: "Package Overview"
vignette: >
  %\VignetteIndexEntry{Package Overview}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

## Purpose

CAISOPEA provides tools for working with CAISO public bid data and related resource-level signals. It helps collect bid data, inspect individual curves, compare anonymous
sequence IDs by feature profiles, and build candidate links between anonymous bid IDs and named resources using outage and curtailment behavior.

The package does not claim perfect deanonymization. Its matching functions produce candidate mappings with scores, confidence values, and detail tables so
the results can be audited.
## Workflow

A typical analysis has five stages:

1. Fetch CAISO OASIS bid files.
2. Filter or load bid data for the market and product of interest.
3. Inspect individual bid curves.
4. Compare anonymous IDs using precomputed feature similarity.
5. Match anonymous IDs to named resources using planned curtailments or full
   outage gaps.

```{r}
#| label: setup
#| message: false
Sys.setenv(
  DUCKPLYR_FALLBACK_COLLECT = "0",
  DUCKPLYR_FALLBACK_AUTOUPLOAD = "0",
  DUCKPLYR_FALLBACK_INFO = "FALSE"
)

library(CAISOPEA)
library(dplyr)

data("caisopea_vignette_sample", package = "CAISOPEA")
```

## Fetch Bid Data

`fetch_oasis_bids()` downloads CAISO public bid data from OASIS and writes one Parquet file per market day. Use `market = "DAM"` for Day-Ahead Market data and `market = "RTM"` for Real-Time Market data.

```{r}
#| eval: false
fetch_oasis_bids(
  start_date = "2025-01-01",
  end_date = "2025-01-07",
  market = "DAM",
  output_dir = "data/dam",
  skip_existing = TRUE
)
```

The return value is a status table with one row per requested date. That makes it easy to see which files were written, skipped, returned no data, or failed after retries.

## Load and Filter Bids

The package functions expect ordinary data frames or compatible lazy tables. For bid-curve analysis, the most commonly used columns are:

- `RESOURCEBID_SEQ`: anonymous bid sequence identifier
- `STARTTIME`: bid operating date or timestamp
- `SCH_BID_XAXISDATA`: bid quantity in MW
- `SCH_BID_Y1AXISDATA`: bid price
- `SELFSCHEDMW`: self-scheduled MW, used by matching functions when computing
  daily maximum MW

```{r}
#| label: sample-tables
lapply(
  caisopea_vignette_sample[
    c(
      "curtailment_dam_data",
      "rtm_data",
      "dam_plot_data",
      "outages",
      "id_attrs"
    )
  ],
  nrow
)
```

For a full local analysis, the same columns can be selected from the large, downloaded Parquet files:

## Plot Bid Curves

`plot_curves()` draws hourly bid curves for one anonymous sequence ID. This is the quickest way to inspect the shape of a resource's submitted bids on a given day or filtered period. The sample includes actual one-day DAM rows from `CAISO_dam_big.parquet`.

```{r}
#| label: overview-plot-curves
#| fig-width: 7
#| fig-height: 4
plot_curves(
  market_data = caisopea_vignette_sample$dam_plot_data,
  resource_id = caisopea_vignette_sample$plot_resource
)
```

The function returns a `ggplot2` object, so it can be printed directly or customized with additional `ggplot2` layers.

## Find Feature-Similar IDs

`find_similar_ids()` uses the package's internal weighted cosine similarity matrix. It is useful when an ID appears to change, split, or have a nearby behavioral peer in the bid-curve feature space.

```{r}
#| label: overview-similar-ids
find_similar_ids(
  id = caisopea_vignette_sample$plot_resource,
  n = 10,
  start_filter = "none"
)
```

`start_filter = "after"` keeps only candidate IDs whose first observed date is after the queried ID's start date. `start_filter = "before"` does the opposite. Those filters are useful when looking for possible ID transitions.

## Match Planned Curtailments

`get_curtailment_matches()` links named resources to anonymous DAM sequence IDs using planned partial curtailments. The idea is that a generator with a known derate before the day-ahead market closes should generally bid up to its available capacity:

$$
\text{expected available MW} =
\text{resource PMAX MW} - \text{curtailment MW}
$$

The function compares that value with each compatible sequence ID's daily maximum bid MW on the same date.

```{r}
#| label: overview-curtailment-match
curtailment_matches <- get_curtailment_matches(
  dam_data = caisopea_vignette_sample$curtailment_dam_data,
  outages = caisopea_vignette_sample$outages,
  id_attrs = caisopea_vignette_sample$id_attrs,
  tolerance = 0.05,
  pmax_tolerance = 0.10,
  min_matches = 2,
  return_detail = TRUE
)

curtailment_matches$final_matches
```

Use `return_detail = TRUE` when developing or auditing results. The returned detail tables show which dates supported each match and how close the observed MW was to the expected available MW.

## Match Full RTM Outages

`get_outage_rtm_matches()` uses a different signal. Full generator outages should create gaps in RTM bids. The function finds bid gaps for anonymous RTM sequence IDs and aligns them with named generator outage windows from the CAISO prior trade day outage data.

```{r}
#| label: overview-outage-match
outage_matches <- get_outage_rtm_matches(
  rtm_data = caisopea_vignette_sample$rtm_data,
  outages = caisopea_vignette_sample$outages,
  generators = caisopea_vignette_sample$final_matches$`RESOURCE ID`,
  edge_minutes = 360,
  min_matches = 1,
  return_detail = TRUE
)

outage_matches$final_matches
```

This compact sample uses a wider edge window so the small slice of actual RTM rows still produces demonstrative matches. For full local studies, start with a tighter `edge_minutes` value and widen only if the outage timestamps and RTM bid interval boundaries are known to be coarser.

## Interpreting Match Output

The matching functions return candidate maps rather than ground truth. Treat the following columns as audit signals:

- `n_matches`: how many dates or outage windows supported the candidate
- `mean_pct_err`: average absolute percentage error in matched capacity
- `score`: combined ranking metric that rewards more matches and lower error
- `confidence`: how clearly the top candidate beat the runner-up for that named
  resource

High confidence is not a substitute for review. Before using a match in downstream analysis, inspect the detail table, check for collisions, and compare the candidate's bid curves against the resource story implied by the outage or curtailment record.
